---
title: "Lake Claremont Report 2023"
output: 
  html_document: 
    toc_depth: 2
    fig_caption: yes
    number_sections: no
    self_contained: no
  word_document:
    toc: yes
    fig_caption: yes
editor_options: 
  markdown: 
    wrap: 80
---

The first code chunk just loads the data and packages, and is the same as the
first R markdown notebook.

```{r load-packages-and-data,include=FALSE}
git <- "https://github.com/Ratey-AtUWA/LC_report_2023/raw/main/"
lcsed23 <- read.csv(paste0(git,"lcsed23.csv"))
lcwater23 <- read.csv(paste0(git,"lcwater23.csv"))
lcwater23$Group <- as.factor(lcwater23$Group)

WQGF <- read.csv("WQ_guide_freshwater_modified.csv")
row.names(WQGF) <- WQGF$Element; row.names(WQGF)[21:22] <- c("P","PO4.P") 
ISQG <- read.csv("ISQG.csv"); row.names(ISQG) <- ISQG$Tox
options("rgdal_show_exportToProj4_warnings"="none")
library(RcmdrMisc)
library(sf)
library(reshape2)
library(maptiles)
library(prettymapr)
library(viridis)       # colorblind-friendly palettes
library(car)
library(diptest)
library(LaplacesDemon)
library(flextable)
  set_flextable_defaults(font.family = "Arial", font.size = 11, 
                       theme_fun = "theme_zebra", padding = 1)
knitr::opts_chunk$set(comment = NA)
```

### Assess variable distributions ± need for transformations

```{r distribution-tests-water, message=FALSE, warning=FALSE, results='hold', echo=FALSE}
# load car package (for powerTransform function) if not loaded yet
require(car)
require(diptest)
require(LaplacesDemon)
require(flextable)

# we're going to change the data frame so let's make a temporary copy
data_0 <- lcwater23
names.of.cols <- names(data_0)

# use column numbers to select variables we want to assess distributions for
# for the Lake Claremont water data, don't use Pb and Sc i.e. c(7:26,28,29,31:34)

colz <- c(7:26,28,29,31:34) ### if soil/sediment data use c(7,8,12:43)
n <- length(colz)

# remove zero or negative values
for(i in colz){
  data_0[which(data_0[,i]<1e-6),i] <- NA
}

# make initial output data frame
transf_results <- data.frame("Variable"=rep(NA,n), "W_orig"=rep(NA,n),
                             "p_orig"=rep(NA,n), "W_log_tr"=rep(NA,n),
                             "p_log_tr"=rep(NA,n), "W_pow_tr"=rep(NA,n),
                             "p_pow_tr"=rep(NA,n), "Pow_term"=rep(NA,n),
                             "diptest_p"=rep(NA,n), "modes"=rep(NA,n))

# start loop that assesses variable distributions and 
# generates matrix of test statistics and p-values

for (i in colz) {
  pt1 <- powerTransform(data_0[, i])
  # generate and print test statistics
  sw0 <- shapiro.test(data_0[, i])
  sw1 <- shapiro.test(log10(data_0[, i]))
  sw2 <- shapiro.test((data_0[, i]) ^ as.vector(pt1$lambda))
  dt0 <- dip.test(log10(data_0[, i]))
  mod0 <- LaplacesDemon::Modes(log10(data_0[, i]))
  transf_results[i-(colz[1]-1),] <- 
    c(names.of.cols[i], signif(sw0$statistic, 3),
      signif(sw0$p.value, 4), signif(sw1$statistic, 3),
      signif(sw1$p.value, 4), signif(sw2$statistic, 3),
      signif(sw2$p.value, 4), signif(as.vector(pt1$lambda), 2),
      signif(dt0$p.value, 4), length(mod0$modes))
}
# output 
for(i in 2:ncol(transf_results)){
  transf_results[,i] <- as.numeric(transf_results[,i])
}
# t0 <- cbind(transf_results[,1], signif(as.numeric(transf_results[,2:ncol(transf_results)],3)))
flextable(transf_results[-c(21,24),]) |>
  set_formatter(p_orig = function(x) sprintf("%.02e",x),
                p_log_tr = function(x) sprintf("%.02e",x),
                p_pow_tr = function(x) sprintf("%.02e",x),
                diptest_p = function(x) sprintf("%.02e",x)) |>
  set_caption(caption="Normality test results for untransformed, log-transformed, and Box-Cox power-transformed variables, plus modality tests, for analysis of water at Lake Claremont. Modality is assessed using log-transformed variables.")

# remove temporary objects to keep R workspace tidy
rm(list=c("data_0","colz","n","names.of.cols","i",
          "pt1","sw0","sw1","sw2","dt0","mod0","transf_results"))
```

# Estimating sulfate/chloride ratio

The calculation is based on the assumptions that:

1. the S measured by ICP-OES is mainly sulfate
2. the ratio of Na to Cl is the same as in seawater 
   (as this ratio is also the same in most rainfall)

In that case, 

- the mass concentration of sulfate is S × 3, since the atomic mass of S is 32, 
  formula mass of SO4 is 96, so 96/32 = 3
- the concentrations (mg/L) of Na and Cl in seawater are Na = 10759, Cl = 19357,
  so Cl = Na × (19357/10759), *i.e*. Cl = Na × 1.799
  (Chester, R. (2009). Marine Geochemistry (2nd ed.). Wiley/Blackwell Science.)
  
$SO_{4}/Cl^{-} = (S \times 3) \div (Na \times 1.799)$

```{r est-sulfate-chloride-ratio, fig.width=7, fig.height=5}
lcwater23$SO4.Cl <- with(lcwater23, (S*3)/(Na*1.799))
par(mar=c(4,4,1,1), mgp=c(2,0.3,0), tcl=0.25, font.lab=2, cex.lab=1.25, las=1)
with(lcwater23, stripchart(SO4.Cl ~ Group, method="jitter", 
                           xlab=expression(bold(paste("SO"[4]^"2\u2212",
                                                      "/Cl"^"\u2212","ratio"))),
                                 ylab = "ENVT4461 Group", 
                           xlim=c(0.13,0.51), pch=1, lwd=2, 
                           col=plasma(nlevels(lcwater23$Group), end=.9)))
abline(v=c(0.14, 0.5), lty=2, col=c("darkmagenta","sienna"))
text(c(0.141,0.498) ,5.5, pos=c(1,3), cex = 0.95, offset = 0.05, 
     col = c("darkmagenta","sienna"),
     labels=c(expression(paste("Mean marine ratio: SO"[4]^"2\u2212",
                             "/Cl"^"\u2212","= 0.14")), 
              expression(paste("Acid sulfate Trigger SO"[4]^"2\u2212",
                               "/Cl"^"\u2212","= 0.5"))), srt=90)

```

Additional exploratory data analysis would include simple maps, scatter plot 
matrices, box plots of variables ~ Group, and so on!
